hormones

### 2019-10-11 analysis/07_hormones.Rmd
now with hormone characterization data

```{r wrangle}
prolactin <- read_excel("../results/Pigeon prolactin concentrations juil 2018.xlsx", sheet = 1)

@@

# create lists of factos for characterization and manipluation studies
charlevels <- c("control", "bldg", "lay", "inc_d3", "inc_d9", "inc_d17", "hatch", "n5", "n9" )
maniplevels <- c("M_inc3", "early", "M_inc9", "M_inc17", "M_n2", "prolong", "extend")
combolevels <- c("control", "bldg", "lay", "inc_d3", "M_inc3",
                  "inc_d9",  "M_inc9",  "inc_d17","M_inc17",  "hatch", "M_n2",
                  "n5", "n9" , "early",  "prolong", "extend")

# rename some of the treatments levels, store in new columns
prolactin <- prolactin %>%
    mutate(sex = fct_recode(Sex,
                            "female" = "f",
                            "male" = "m"),
           treatment = fct_recode(Treatment,
                            "hatch" = "Hatch",
                            "inc_d17" = "Inc_d17",
                            "inc_d3" = "Inc_d3",
                            "inc_d9" = "Inc_d9",
                            "M_inc9" = "M_Inc9",
                            "M_inc3" = "M_Inc3",
                            "early" = "M_Inc8",
                            "early" = "M_inc8",
                            "M_inc17" = "M_Inc17",
                            "M_n2" = "M_hatch",
                            "control" = "baseline",
                            "n5" = "N5", 
                            "n9" = "N9"),
           study = fct_collapse(treatment,
                                 characterization = charlevels,
                                 manipulation = maniplevels))

prolactin$treatment <- factor(prolactin$treatment, levels = combolevels)
```

```{r wranglesteroids}
PETC <- read_excel("../results/parental_care_hormone_RIA_data_master.xlsx", sheet = 2)

@@

PETC <- PETC %>%
    mutate(sex = fct_recode(sex,
                            "female" = "f",
                            "male" = "m"),
           study = fct_collapse(stage,
                                 characterization = charlevels,
                                 manipulation = maniplevels))
PETC$treatment <- factor(PETC$stage, levels = combolevels)

@@

```{r prolactin}
prolactin %>%
  select(treatment, `Prolactin ng/mL`, Sex, study) %>% drop_na() %>%
ggplot(aes(x = treatment, y = `Prolactin ng/mL`, fill = Sex)) +
  geom_boxplot() + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "bottom") +
  facet_wrap(~study, scales = "free_x") +
  labs(y = "Prolactin (ng/mL)")
```

```{r testosterone}

PETC %>% 
  filter(hormone == "testosterone", sex == "male")  %>% 
ggplot(aes(x = treatment, y = plasma_conc, fill = sex)) +
  geom_boxplot() + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "bottom") +
  facet_wrap(~study, scales = "free_x") +
  labs(y = "Testosterone") +
  scale_fill_manual(values = c("#00BFC4"))
```

```{r estradiol}
PETC %>% 
  filter(hormone == "estradiol", sex == "female")  %>% 
ggplot(aes(x = treatment, y = plasma_conc, fill = sex)) +
  geom_boxplot() + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "bottom") +
  facet_wrap(~study, scales = "free_x") +
  labs(y = "Estradiol (ng/mL)")

@@

```{r progesterone}
PETC %>% 
  filter(hormone == "progesterone")  %>% 
ggplot(aes(x = treatment, y = plasma_conc, fill = sex)) +
  geom_boxplot() + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "bottom") +
  facet_wrap(~study, scales = "free_x") +
  labs(y = "Progesterone (ng/mL)") +
  ylim(c(0,6))

```{r cort}

PETC %>% 
  filter(hormone == "cort")  %>% 
ggplot(aes(x = treatment, y = plasma_conc, fill = sex)) +
  geom_boxplot() + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "bottom") +
  facet_wrap(~study, scales = "free_x") +
  labs(y = "Corticosterone (ng/mL)") +
  ylim(c(0,10))

```{r prolactin-char}
characterization <- prolactin %>% filter(treatment %in% charlevels)   %>%  droplevels()

mod_prolactin <-  lm(data = characterization, `Prolactin ng/mL` ~ treatment + sex)
mod_prolactin

grid <- characterization %>%
  data_grid(treatment, sex) %>%
  add_predictions(mod_prolactin)

grid

df <- data.frame(
  x = c(1,2,3,4,5,6,7,8,9),
  y = c(120,120,120,120,120,120,120,120,120),
  sex = c("female", "female", "female", "female", "female", "female","female", "female", "female"),
  text = c("AB",  "A", "A" , "AB", "AB", "BCD", "D", "CD", "BC")
)

a <- ggplot(characterization, aes(x = treatment, y = `Prolactin ng/mL`, fill = sex)) +
  geom_boxplot() + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "bottom",
        legend.title = element_blank()) +
  labs(subtitle = "Characterization", y = "Prolactin (ng/mL)", x = NULL) +
  #geom_text(data = df, aes(x,y, label = text), color = "darkgrey") +
      ylim(0,120) 
a

b <- ggplot(grid, aes(x = treatment, y = pred, color = sex)) +
  geom_point() + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "bottom",
        legend.title = element_blank()) +
  labs(subtitle = "Model prediction ",
       y = NULL, x = NULL) +
  geom_smooth(se = FALSE, aes(x = as.numeric(treatment)))  +
  ylim(0,120)

plot_grid(a,b)
```

```{r aov}
aov1 <- aov(data = characterization, `Prolactin ng/mL` ~ treatment + sex)
aov1
summary(aov1)
TukeyHSD(aov1, which = "treatment")
```

```{r prolactin-manip, fig.width=8} 
manipulation <- prolactin %>% 
  filter(treatment %in% maniplevels) %>%  droplevels() %>% 
  mutate(result = fct_collapse(treatment,
                                "early hatch" = c("early"),
                                "offspring removed" = c("M_inc3",  "M_inc9", "M_inc17", "M_n2"),
                               "late incubation" = c("prolong"),
                               "late hatch" = c("extend")))

e <- manipulation %>%
  ggplot( aes(x = treatment, y = `Prolactin ng/mL`)) +
    geom_boxplot(aes(fill = sex)) + 
    theme(axis.text.x = element_text(angle = 45, hjust = 1),
          legend.position = "bottom",
          legend.title = element_blank()) +
    ylim(0,125) +
    labs(subtitle = "Manipulation",
       y = "Prolactin (ng/mL)", x = NULL)
e

plot_grid(a,e, nrow = 1)

f <-  prolactin %>%
  filter(treatment %in% c("control", "bldg", "inc_d3", "M_inc3",
                          "inc_d9",  "M_inc9", "inc_d17","M_inc17",
                          "hatch", "M_n2")) %>%
ggplot(aes(x = treatment, y = `Prolactin ng/mL`, fill = Sex)) +
  geom_boxplot(aes(alpha = study)) + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "none") +
  labs(subtitle = "Offspring removed",
       x = NULL) +
  facet_wrap(~sex) +
  annotate("rect", xmin = 2.5, xmax = 4.5, ymin = 0, ymax = 120, alpha = .2) +
    #annotate("rect", xmin = 4.5, xmax = 6.5, ymin = 0, ymax = 120, alpha = .3) +
    annotate("rect", xmin = 6.5, xmax = 8.5, ymin = 0, ymax = 120, alpha = .2) +
    #annotate("rect", xmin = 8.5, xmax = 10.5, ymin = 0, ymax = 120, alpha = .3) +
  ylim(0,120) +
  scale_alpha_manual(values = c(1,0.5))

g <-  prolactin %>%
  filter(treatment %in% c("early","prolong", "extend",
                          "inc_d9","inc_d17","hatch", "n5" )) %>% droplevels() %>%
ggplot(aes(x = treatment, y = `Prolactin ng/mL`, fill = Sex)) +
  geom_boxplot(aes(alpha = study)) + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "none") +
  labs(subtitle = "Timing of transition altered",
       x = NULL, y = NULL) +
  facet_wrap(~sex) +
    annotate("rect", xmin = 4.5, xmax = 7.5, ymin = 0, ymax = 120, alpha = .3) +
  ylim(0,120) +
  scale_alpha_manual(values = c(1,0.5))

plot_grid(f,g, nrow=2, rel_widths = c(.625,.375))

```{r fortwitter, fig.height= 3, fig.width=5.5}
ggplot(grid, aes(x = treatment, y = pred, color = sex)) +
  geom_point() + 
  theme(axis.text.x=element_blank(),
        axis.text.y=element_blank(),
        legend.position = c(0.75,0.3),
        legend.title = element_blank()) +
  labs(y = "Gene expression", x = "Time") +
  geom_smooth(se = FALSE, aes(x = as.numeric(treatment)))  
```

```{r}
write.csv(prolactin, "../results/prolactin.csv", row.names = F)
```

>>>>

```{r clean}
charlevels <- c("control", "bldg", "lay", "inc_d3", "inc_d9", "inc_d17", "hatch", "n5", "n9" )
removelevels <- c("M_inc3",  "M_inc9", "M_inc17", "M_n2")
maniplevels <- c("early", "prolong", "extend")
combolevels <- c("control", "bldg", "lay", "inc_d3", "M_inc3",
                  "inc_d9",  "M_inc9", "early",  "inc_d17","M_inc17",  "hatch",
                 "prolong", "extend", "M_n2",
                  "n5", "n9" )
```

```{r wrangle-prolactin}
prolactin <- read_excel("../results/Pigeon prolactin concentrations juil 2018.xlsx", sheet = 1)

@@

# create lists of factos for characterization and manipluation studies

# rename some of the treatments levels, store in new columns
prolactin <- prolactin %>%
    mutate(sex = fct_recode(Sex,
                            "female" = "f",
                            "male" = "m"),
           treatment = fct_recode(Treatment,
                            "hatch" = "Hatch",
                            "inc_d17" = "Inc_d17",
                            "inc_d3" = "Inc_d3",
                            "inc_d9" = "Inc_d9",
                            "M_inc9" = "M_Inc9",
                            "M_inc3" = "M_Inc3",
                            "early" = "M_Inc8",
                            "early" = "M_inc8",
                            "M_inc17" = "M_Inc17",
                            "M_n2" = "M_hatch",
                            "control" = "baseline",
                            "n5" = "N5", 
                            "n9" = "N9"),
           study = fct_collapse(treatment,
                                 characterization = charlevels,
                                 manipulation = maniplevels,
                                removal = removelevels))

colnames(prolactin)[colnames(prolactin)=="Prolactin ng/mL"] <- "plasma_conc"
prolactin$hormone <- "prolactin"
prolactin <- prolactin %>% select(study, treatment, sex, hormone, plasma_conc)   %>% drop_na()

summary(prolactin)
```

```{r wrangle-steroids}
PETC <- read_excel("../results/parental_care_hormone_RIA_data_master.xlsx", sheet = 2)

@@

PETC <- PETC %>%
    mutate(sex = fct_recode(sex,
                            "female" = "f",
                            "male" = "m"),
           study = fct_collapse(stage,
                                 characterization = charlevels,
                                manipulation = maniplevels,
                                removal = removelevels)) %>% 
  drop_na()
colnames(PETC)[colnames(PETC)=="stage"] <- "treatment"
PETC <- PETC %>% select(study, treatment, sex, hormone, plasma_conc)
summary(PETC)
```

```{r combine-clean-split}
hormones <- rbind(prolactin, PETC)
hormones$treatment <- factor(hormones$treatment, levels = combolevels)

hormones$okay <- ifelse(hormones$hormone == "cort" & hormones$plasma_conc > 30, "bad",
                    ifelse(hormones$hormone == "progesterone" & hormones$plasma_conc > 5, "bad", 
                           ifelse(hormones$hormone == "prolactin" & hormones$plasma_conc > 150, "bad", 
                            ifelse(hormones$hormone == "prolactin" & hormones$treatment == "control", "bad", 
                        ifelse(hormones$hormone == "testosterone" & hormones$sex == "female", "bad",
                               ifelse(hormones$hormone == "estradiol" & hormones$sex == "male", "bad", "okay"))))))
hormones <- hormones %>% filter(okay == "okay") %>% droplevels()

prl.char <- hormones %>% filter(hormone == "prolactin", study == "characterization")   %>%  droplevels()
test.char <- hormones %>% filter(hormone == "testosterone", treatment %in% charlevels)   %>%  droplevels()
est.char <- hormones %>% filter(hormone == "estradiol", treatment %in% charlevels)   %>%  droplevels()
prog.char <- hormones %>% filter(hormone == "progesterone", treatment %in% charlevels)   %>%  droplevels()
cort.char <- hormones %>% filter(hormone == "cort", treatment %in% charlevels)   %>%  droplevels()

@@

```{r characterization-hormones}

hormonecharplot <- function(myhormone, myylab){

  mycolors <- c("female" = "#F8766D", "male" = "#00BFC4")

  hormones %>% 
    filter(study == "characterization",
           hormone %in% c(myhormone))  %>% 
  ggplot(aes(x = treatment, y = plasma_conc, fill = sex)) +
    geom_boxplot() + 
    theme(axis.text.x = element_text(angle = 45, hjust = 1),
          legend.position = "none") +
    scale_fill_manual(values = mycolors) +
    labs(y = myylab, x = NULL) 
}

a <- hormonecharplot("estradiol", "E (ng/mL)")
b <- hormonecharplot("testosterone", "T (ng/mL)")
c <- hormonecharplot("cort", "CORT (ng/mL)")
d <- hormonecharplot("progesterone", "PROG (ng/mL)")

e <- hormonecharplot("prolactin", "PRL (ng/mL)")

# model and plot prolactin data
mod_prolactin <-  lm(data = prl.char,  plasma_conc ~ treatment + sex)
grid <- prl.char %>%
  data_grid(treatment, sex) %>%
  add_predictions(mod_prolactin)

f <- ggplot(grid, aes(x = treatment, y = pred, color = sex)) +
  geom_point() + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "none",
        legend.title = element_blank()) +
  labs(y = "predicted PRL (ng/mL)", x = NULL) +
  geom_smooth(se = F, aes(x = as.numeric(treatment)))

plot_grid(e,f,  rel_widths = c(0.65,0.35))

plot_grid(d + theme(axis.text.x=element_blank()),
          a + theme(axis.text.x=element_blank()),
          c, b, 
          rel_widths = c(0.65,0.35),
          rel_heights = c(0.45,0.55), ncol = 2)

@@

```{r aov}
aov1 <- aov(data = prl.char, plasma_conc ~ treatment + sex)
aov1
summary(aov1)
TukeyHSD(aov1, which = "treatment")

aov2 <- aov(data = prog.char, plasma_conc ~ treatment + sex)
aov2
summary(aov2)

aov3 <- aov(data = cort.char, plasma_conc ~ treatment + sex)
aov3
summary(aov3)
TukeyHSD(aov3, which = "treatment")

aov4 <- aov(data = est.char, plasma_conc ~ treatment )
aov4
summary(aov4)
TukeyHSD(aov4, which = "treatment")

aov5 <- aov(data = test.char, plasma_conc ~ treatment )
aov5
summary(aov5)
```

```{r}
write.csv(hormones, "../results/hormones.csv", row.names = F)
```
